Identification of rice genes which respond to saline stress from co-expression network analysis

### 2020-09-29 camila/Report/article_CR.tex
new paper draft

%--------------------------------------------------------------------------
\title{Identification of rice genes which respond to saline stress from co-expression networks analysis}
\author{Camila Riccio Rengifo\\
  \small Pontificia Universidad Javeriana Cali\\
}

@@

\subsection{Co-expression network}
A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of edges or links that connect the vertices. In a gene co-expression network each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns. These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\

When the network is simple and unweighted, it can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$. This adjacency matrix is symmetric, with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere.\\

@@

Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}. HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. The similarity between two links $e_{ik}$ and $e_{jk}$ is computed using the Jaccard index 
\begin{equation*}
\label{eq:jaccard}
S(e_{ik},e_{jk}) = \frac{\vert n_+(i) \cap n_+(j) \vert}{\vert n_+(i) \cup n_+(j) \vert},
\end{equation*}
where $n_+(i)$ denotes the set of node $i$ and its neighbors.\\

With this similarity, the method uses single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities. Hierarchical clustering methods repeatedly merge groups until all elements are members of a single cluster. To find meaningful communities it is crucial to know where to partition the dendrogram. Thus, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities that measures the quality of a link partition. The partition density has a single global maximum along the dendrogram in almost all cases, because the value is just the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link). $D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree. If a link community is less dense than a tree (when the community subgraph has disconnected components), then that community will give a negative contribution to $D$. Computing $D$ at each level of the link dendrogram allows to pick the best level to cut (although meaningful structure exists above and below that threshold). This way, the final output corresponds to node clusters, where each node can participate in multiple communities.\\

\subsection{Least Absolute Shrinkage Selector Operator}
Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients. Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ m $ ($ n \gg m $). Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:

@@

Or equivalently minimizing:
\begin{equation}
\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} + \lambda \sum_{j=1}^n\abs{\beta_j}
\end{equation}
being $ s $, $ \lambda \geq 0 $ the respective penalty parameters for complexity.\\

@@

Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice. Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error. With that selected $\lambda$ value, the model is adjusted again, using all the observations.\\

@@

This methodology uses as input data RNA-seq read counts, representing gene expression levels. More precisely, $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates. 
This raw data can be represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$. 
In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes. The phenotypic data can be seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second under treatment condition.\\

@@

The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}. From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$. Next, remove genes exhibiting low variance or low expression, getting a new number of $n_1 \leq n_0$ genes. Separate control and treatment data into two matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively. Each matrix entry, $c_{ij}$ for control and $t_{ij}$ for treatment, represent the normalized expression level of gene $i$ in accession $j$.\\

Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios. Matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Repeat this process on the phenotypic data separating the control and treatment data into matrices $P_c$ and $P_t$ of dimensions $m \times p$. Calculate the corresponding logarithmic ratios, obtaining the matrix $P_\ell \in \mathbb{R}^{m \times p}$. Finally, filter $L_0$ matrix by removing rows (genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\

@@

The Log fold change matrix $L_1$ is used to build the co-expression network following the first steps of the WGCNA methodology. First, measure the level of concordance between gene differential expression profiles across samples. Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\

Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). \\

Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$. To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values. 
Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\

@@

Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules. The idea is to cluster genes with similar differential expression change patterns. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $\hat{A}$, from the co-expression network $G$, is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\

@@

The HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community. This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of cluster $u$.\\

@@

To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach. Each module can be represented by a eigengene, which is defined as the first principal component of such module. A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change Matrix $L_1$ and the affiliation matrix $F$. Given a module $u$, use the affiliation matrix to identify the genes belonging to $u$ and then select the corresponding rows of the matrix $L_1$ to compute the first principal component of $u$. Each principal component would be a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\

@@

These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO). LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients. This technique allows selection of the most relevant variables. In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable. \\

The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.

@@

For the genes found, first identify the differential expressed ones, that is genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample. This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\

You also can perform a functional category enrichment searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}. This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\

Those named genes (with GO annotations), can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}. The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases. This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.\\

\section{Case study}
RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates. On the other hand, we use $p=3 $ phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. This traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\

\subsection{Data pre-processing}
The DESeq normalization was applied to the raw data and the biological replicates were averaged. Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed. Genes with low expression were also removed, corresponding to those having more than $80\%$ samples with values smaller than $10$. After this filtering process, $n_1 = 8928$ genes were kept.\\

From the compiled matrix $L_0$ we remove genes whose difference between upper quantile and lower quantile is greater than $0.25$. So the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes. We also compute the logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits.\\

\subsection{Co-expression network construction}
We used the Log Fold Change matrix $L_1$ to compute the corresponding similarity matrix. 
We found that for this network, the smallest integer such that the $R^2 \geq 0.8 $ is $\beta=3$. Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.

@@

The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes (nodes) and $|E| = 39850128$ edges. We choose $0.2$ as cutoff value, keeping only the connections above this threshold and removing the isolated nodes, so the new adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\

@@

\subsection{Modules association to phenotypic trits}
The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass. Figure \ref{fig:pdata} shows that there are significant differences in the values of these phenotypic traits between stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.

@@

Using the affiliation matrix $F$ derived from HLC output and the Log Fold Change matrix $L_1$, we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules.
We apply the LASSO technique using each of the phenotypic traits as the outcome variable, one at a time. As shown in Figure~\ref{fig:cross-val} we perform cross-validation for each phenotypical trait, in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\

@@

We finally adjust the three models using all observations and the corresponding $\lambda$ values. As result, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot $K$ content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\

@@

On the one hand, SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars. On the other hand plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding.This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism.\\

Previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars. However, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.

@@

The methodology described in this manuscript provide an aproach to discover key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules. Similar approximations have been proposed~(\cite{langfelder2008wgcna, du2019network}), but they only focus on disjoint (non-overlapping) communities. However, it is well known that communities in real networks are overlapping. \\

>>>>

%--------------------------------------------------------------------------
\title{Identification of rice genes which respond to saline stress from co-expression network analysis}
\author{Camila Riccio Rengifo\\
  \small Pontificia Universidad Javeriana Cali\\
}

@@

\subsection{Co-expression network}
A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of \textit{vertices} or \textit{nodes} and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of \textit{edges} or \textit{links} that connect vertices. In a gene co-expression network, each node corresponds to a gene. A pair of genes is connected if they show similar differential expression patterns. A simple and unweighted network can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$ that is symmetric  with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere. Co-expression networks are of biological interest because the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\

@@

The Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}. The HLC approach represents communities as groups of links (rather than nodes), each node inherits all memberships of its links and can thus belong to multiple, overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. The similarity between two links $e_{ik}$ and $e_{jk}$ is computed using the Jaccard index 
\begin{equation}
\label{eq:jaccard}
S(e_{ik},e_{jk}) = \frac{\vert n(i) \cap n(j) \vert}{\vert n(i) \cup n(j) \vert},
\end{equation}
where $n(i)$ denotes the set corresponding to node $i$ and its neighbors. The algorithm uses single-linkage hierarchical clustering to build a dendrogram in which each leaf is a link from the original network and branches represent link communities. Hierarchical clustering algorithms repeatedly merge groups until all elements are members of a single cluster.\\

For the purpose of finding meaningful communities it is crucial to know where to partition the dendrogram. In this case, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities measuring the quality of a link partition. The partition density has a single global maximum along the dendrogram in almost all cases, because the value is the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link). It is the case that $D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree. If a community is less dense than a tree (when the community subgraph has  disconnected components), then that community will give a negative contribution to $D $, which can take values less than zero.The minimum density inside a community is $-2/3$,  given by one community of two disconnected edges. Since $D$ is the average of the intra-community density, there is a lower bound of $D=-2/3$. Computing $D$ at each level of the link dendrogram can help the purpose of picking the best level to cut (although meaningful structure could exist above or below the threshold). The output of the process is a set of node clusters, where each node can participate in multiple communities.\\

\subsection{Least Absolute Shrinkage Selector Operator}
The Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique. It combines a regression model, with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients. In other words, Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector. It can be especially useful to solve problems where the number of variables (e.g., genes) $ n $ is much greater than the number of samples $ m $ (i.e., $ n \gg m $).\\

Consider a dataset consisting of $m$ samples, each of which consists of $n$ covariates and a single outcome. Let $y_i$ be the outcome and $x_i := (x_1,...,x_n)$ be the covariate vector for the $i-th$ sample. Then, the objective of LASSO is to solve

@@

Or, equivalently, in the so-called Lagrangian form, it minimizes

\begin{equation}
\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} + \lambda \sum_{j=1}^n\abs{\beta_j}
\end{equation}

where $s$ is the regularization penalty and $\lambda \geq 0$ is the corresponding Lagrange multiplier.\\

@@

Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice. Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error.\\

@@

The proposed methodology uses RNA-seq read counts, representing gene expression levels, as input data. More precisely, it uses $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates. This raw data is represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$. In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the approach also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes. The phenotypic data is seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second one under treatment condition. The whole methodology can be followed in the flow chart of Figure~\ref{fig:flow_chart}.\\

% flow chart
\begin{figure}[]
  \centering
    \includegraphics[clip,width=1\textwidth]{Figures/flow_chart.pdf}
  \caption{Proposed methodology flow chart, showing the five macro-processes: \textbf{A}.~Data pre-processing, \textbf{B}.~Co-expression network contruction, \textbf{C}.~Co-expression module identification, \textbf{D}.~Modules association to phenotypic traits, and \textbf{E}.~Genes enrichment.}
  \label{fig:flow_chart}
\end{figure}

@@

The RNA-seq data cannot be directly interpreted. Therefore, a normalization process is applied to deal with the problem of some biases affecting the quantification results. In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}. The normalized data is represented as a matrix $D_1 \in \mathbb{R}^{n_0 \times 2mr}$. The biological replicates of each genotype are averaged and the resulting data is represented as a matrix $D_2 \in \mathbb{R}^{n_0 \times 2m}$. Next, the genes exhibiting low variance or low expression are removed from $D_2$, thus identifying a subset of size $n_1 \leq n_0$ of the original genes. The control and treatment data is separated into the matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively. The matrix entries $c_{ij}$ in $C$ and $t_{ij}$ in $T$ represent, respectively, the normalized expression level of gene $i$ in accession $j$. Control and treatment data is also separated from phenotypic data $P$, obtaining the $P_c$ and $P_t$ matrices, of dimensions $m \times p$.\\

%From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$. Next, remove genes exhibiting low variance or low expression, getting a new number of $n_1 \leq n_0$ genes. Separate control and treatment data into two matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively. Each matrix entry, $c_{ij}$ for control and $t_{ij}$ for treatment, represent the normalized expression level of gene $i$ in accession $j$.\\

At this point, the changes in expression levels and phenotypic values, between control and treatment conditions, are measured in terms of logarithmic ratios. In the case of expression levels, the log ratios are represented in a matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, where $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Similarly, the log ratios of the phenotypic data are computed and represented in a matrix $P_\ell \in \mathbb{R}^{m \times p}$.\\

The final step of the data pre-processing is to filter $L_0$ by removing rows (e.g., genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\

%Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios. Matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Repeat this process on the phenotypic data separating the control and treatment data into matrices $P_c$ and $P_t$ of dimensions $m \times p$. Calculate the corresponding logarithmic ratios, obtaining the matrix $P_\ell \in \mathbb{R}^{m \times p}$. Finally, filter $L_0$ matrix by removing rows (genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\

@@

The Log fold change matrix $L_1$ is used to build the co-expression network following the first two steps of the WGCNA methodology~\cite{langfelder2008wgcna}. First, the level of concordance between gene differential expression profiles across samples is measured. To this end, the absolute value of the Pearson correlation coefficient is used as the similarity measure between genes and the resulting values are stored in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\

And second, the matrix $S$ is transformed into and adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). \\

%Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). \\

@@

Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules. The idea is to cluster genes with similar differential expression change patterns. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $A$, is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\

@@

As a previous step, matrix $A$ has to be modified, in order to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$. To this end, the Pearson Correlation Coefficient (PCC) cutoff is determined using the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values. Above the cutoff, the entries of matrix $A$ become $1$, and below the cutoff $A$ values becomes zero.\\

%To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values. Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\

Now, the HLC algorithm can be applied. It organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community. This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of cluster $u$.\\

@@

To identify the most relevant groups (modules) of genes, associated with the phenotypic response to a specific treatment in an organism, the proposed methodology uses a LASSO based approach. Each module is represented by a eigengene, which is defined as the first principal component of such module. A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change Matrix $L_1$ and the affiliation matrix $F$. Given a module $u$, the affiliation matrix is used to identify the genes belonging to $u$ and then the corresponding rows of the matrix $L_1$ are selected to compute the first principal component of $u$. Each principal component becomes a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\

@@

These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO).
%LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients. 
This technique allows selection of the most relevant variables. In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable at a time.\\

The output after applying LASSO is a set of modules for each phenotypic trait. The union of genes belonging to the selected modules are the target genes for downstream analysis.\\

@@

The objective of this part of the analysis is to characterize the genes found with complementary information, helping to elucidate their possible behavior and role in the response to the studied treatment.\\

A simple analysis to made with the selected genes, using the previous calculations, is to identify the differentially expressed ones. That is, identify genes showing an absolute value of the log fold change greater than 2 ($|\ell_{ij}|\geq 2$) for at least one sample. This represents genes whose level of expression is quadrupled (up or down) from control to treatment condition, suggesting that those genes are strong candidates as treatment responsive genes.\\

A widley used approach in this step is to perform a functional category enrichment, searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}. This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\

Addicionally, QuickG0 can be used to identify genes with reported protein products, which can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}. The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases. This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.\\

% Otras caracteristicas pueden buscarse en otras bases de datos o hacer una busqueda más exahustiva en articulos de estudios previos sobre estos genes.

\section{Case study}
The RNA-seq data was obtained from GEO database \cite{GEOAcces90:online} (accession number GSE98455). This data corresponds to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates. A total of $p=3 $ phenotypic traits are used: shoot $K^+$ content, root biomass and shoot biomass. These traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\

\subsection{Data pre-processing}
DESeq2 normalization was applied to the raw data and the biological replicates were averaged. Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed from the normalized data. Genes with low expression, corresponding to those having more than $80\%$ samples with values smaller than $10$, were also removed. A total of $n_1 = 8928$ genes were kept after this filtering process.\\

From the Log Fold Change matrix $L_0$, genes whose difference between upper quantile and lower quantile was greater than $0.25$, were removed. Therefore, the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes. The logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits, was also computed.\\

\subsection{Co-expression network construction}
The Log Fold Change matrix $L_1$ is used to compute the corresponding similarity matrix. 
For this network, it was observed that $\beta=3$ is the smallest integer such that the $R^2 \geq 0.8 $. Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right), which is the degree distribution of a scale-free network with $R^2 = 0.8$ with $\beta = 3$.

@@

The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes and $|E| = 39850128$ edges. The cutoff value was set to $0.2$, thus keeping only the connections above this threshold and removing the isolated nodes. The resulting adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\

@@

\subsection{Modules association to phenotypic traits}
The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there are significant differences in the values of these phenotypic traits between stress and control conditions. This supports the working hypothesis that these three variables represent tolerance-associated traits in rice under salt stress.

@@

Using the affiliation matrix $F$ derived from the HLC output and the Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the egengene for each of the $c = 5143$ modules. The LASSO technique was applied by using each of the phenotypic traits as the outcome variable, one at a time. As shown in Figure~\ref{fig:cross-val} cross-validation was performed for each phenotypical trait in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\

%we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules. We apply the LASSO technique using each of the phenotypic traits as the outcome variable, one at a time. As shown in Figure~\ref{fig:cross-val} we perform cross-validation for each phenotypical trait, in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\

@@

FInally, three LASSO models were adjusted by using the corresponding $\lambda$ and phenotypical data with the egengens of matrix $M$. As result, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot $K$ content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\

@@

Uniprot database~\cite{uniprot2018uniprot} reports, on the one hand, that SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars. On the other hand, is reported that plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding. This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism. Aditionally, previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars. However, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.

@@

This manuscript provides a detailed description of a network analysis-based methodology to 
%The methodology described in this manuscript provides network analysis-based approach to
the discovery of key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules.\\

The proposed methodology follows the principles of WGCNA. Starts from the level of thousands of genes, identifies biologically interesting gene modules, and identifies key genes in the treatment pathways for further validation. In WGCNA the whole workflow is structured in a modular way. The proposed methodology takes advantage of this property modifying and exploring new techniques inside some processes.\\

An important modification to the WGCNA methodology is in the pre-processing of the input data. Our approach is designed to integrate expression data measured under two different conditions (control and treatment), instead of working with them independently or considering a single condition. For this purpose, an approach similar to that proposed in~\cite{du2019network} is used, where the control and treatment data are compiled in a single matrix using the log fold change measure. Thus, the input to construct the co-expression network is not the expression data but the changes in the expression levels from one condition to the other, allowing to capture the signal of changes caused by the treatment.\\

The most innovative feature of the proposed methodology is the module detection technique. 
The co-expression network is calculated as in WGCNA until the scale-free network is obtained. In our approach, this network is used to apply the HLC algorithm, which is capable of detecting overlapping communities. The usual approximation and its different variations, only focuses on disjoint (non-overlapping) communities, but it is well known that communities in real networks are overlapping. Thus the approach presented in this work can be seen as a generalization of the previous approaches.\\

%Similar approximations have been proposed in, e.g., \cite{langfelder2008wgcna} and \cite{du2019network}, but they only focus on disjoint (non-overlapping) communities. It is well known that communities in real networks are overlapping, and thus the approach presented in this work can be seen as a generalization of the previous approaches.\\

% La tercera modificación es incluir LASSO como método para relacionar los datos phenotypicos con los transcriptomicos.
